Enforcing idempotency in neural networks

### 2025-01-30 paper.tex
Before changes suggested by Jamie

% Wider context -> Idempotent networks
Using neural networks as data augmentation tools is becoming more widespread in areas such as signal processing and generative artificial intelligence. In particular, networks of the form ${f: X \to X}$, mapping data within the same space $X$, are frequently used in image augmentation \cite{lu-image-aug}, video generation \cite{ma-compression, liu-gans}, sorting algorithms \cite{tambouratzis-sorting}, compression algorithms \cite{namphol-compression, liu-gans}, image denoising \cite{mao-deblurring, ilesanmi-denoising, liu-gans}, and image generation \cite{liu-gans}, among others. While there is a variety of such transformation tasks, some of these can be considered \textit{intrinsically idempotent} in the sense that any function which carries out the transformation is necessarily an idempotent operator. An idempotent operation is one which can be applied iteratively with no effect beyond the first application. For instance, a sorting transformation is intrinsically idempotent since sorting an already sorted data structure is redundant and such a sorting is often deterministic. On the other hand, image transformations might not always be idempotent: rotating an image by 90$^{\circ}$ twice is not the same as rotating it once, for instance. As such, some transformations are intrinsically idempotent and admit \textit{only} idempotent solutions while other tasks are not and admit \textit{no} idempotent solutions. This work, however, is concerned with a class of data transformation tasks which has both idempotent and non-idempotent solutions and where idempotency might be a desirable property. For example, in section \ref{sec:experiment} we study idempotency in generative networks where it is the formal requirement of one-step inference, but also denoising and image augmentation (e.g., application of effect-filters) are examples of tasks where idempotent solutions may be desirable \cite{mao-deblurring,liu-gans}. Since solutions are not intrinsically idempotent in this class, we explore actively enforcing idempotency as a component of the loss function used in training.

% What is an idempotent network even?
In this paper we are primarily concerned with networks $f_{\vtheta}: \mathbb{R}^n \to \mathbb{R}^n$, where $\vtheta$ is a collection of weight parameters. The condition that $f_{\theta}$ is idempotent is
%
\begin{equation}
    f_{\vtheta}(\vx) = f_{\vtheta}(f_{\vtheta}(\vx))
    \label{eq:idem}
\end{equation}
%
for all $\vx \in \mathbb{R}^n$. If $f_{\vtheta}(\vx) = \vW \vx$ (a single-layer, fully-connected network with no bias and the identity activation function) where $\vW \in \mathbb{R}^{n \times n}$ is the weight matrix, then condition \ref{eq:idem} reduces to the familiar notion from linear algebra where $\vW = \vW^2$ and eigenvalues of $\vW$ are either 0 or 1. Condition \ref{eq:idem} also gives the correct notion for non-linear networks acting as idempotent projections, and can be optimized using a simple mean-squared error loss,
%
\begin{align}
    \mathcal{L}_\text{idem}(\vx) = \frac{1}{m} \sum_{i = 1}^m \left(f_{\vtheta}(f_{\vtheta}(\vx)) - f_{\vtheta}\big(\vx \big)\right)^2,
    \label{eq:idem-loss}
\end{align}
%
where $\vx \in \mathbb{R}^{n}$. As we show in section \ref{sec:experiment}, minimizing this loss using canonical gradient descent can yield relatively poor improvement in the idempotent loss. Additionally, due to the higher-order application of $f_{\vtheta}$ the number of terms in the gradient $\nabla_{\vtheta} \mathcal{L}_{\text{idem}}$ grows exponentially in the number of layers, making the approach computationally expensive for certain architectures.

% What are we proposing in this paper?
In this work, we propose an alternative method for training neural networks to satisfy condition \ref{eq:idem}. Using ideas from Perturbation Theory \cite{intro-pertub-theory} we derive a function $g$ which solves $\vK' = g(\vK)$ such that if $\vK \in \mathbb{R}^{n \times n}$ is an ``almost'' idempotent matrix, then $\vK' \in \mathbb{R}^{n \times n}$ is perfectly idempotent (i.e., $(\vK')^2 = \vK'$). In this work, we focus on one such function:
%
\begin{equation}
    g(\vK) = 3 \vK^2 - 2 \vK^3
    \label{eq:g}
\end{equation}
%
Although we assume $\vK$ is close to idempotent, we show that in practice $g$ can be used to derive matrices which are within machine precision of perfect idempotency even when the input matrix $\vK$ is relatively far from idempotent. At a high level, this process is based on a recurrence relation ${\vK' \leftarrow (1 - \gamma)\vK + \gamma g(\vK)}$, taking small $\gamma$-sized steps in the direction of $g(\vK)$. While this recurrence relation derives idempotent matrices -- and can therefore be used to train single-layer networks with identity activations to be idempotent -- we also give a more general application of Eq. \ref{eq:g} as a minor modification of the backpropagation algorithm, yielding an architecture agnostic and efficient algorithm for finding idempotent networks.

% Outline of each section of the paper
In section \ref{sec:method} we give a detailed description of the method used to derive Eq. \ref{eq:g} and alternative solutions. We also show that while there exists non-idempotent fixed points to Eq. \ref{eq:g}, these points are repelling under the recurrence relation ${\vK' \leftarrow (1 - \gamma)\vK + \gamma g(\vK)}$ for $0 \leq \gamma < 1$, giving credence to the use of such a recurrence relation in practice. Finally, in section \ref{sec:method} we derive a full training scheme for training arbitrary neural network architectures of the form $f_{\vtheta}: \mathbb{R}^n \to \mathbb{R}^n$. In section \ref{sec:experiment}, we present experimental data for a variety of fully-connected network architectures, showing that our method outperforms ordinary backpropagation under varied conditions. We also replicate the results of \cite{shocher-ign} by applying our method on a U-net style DCGAN network to successfully create a generative network for the MNIST dataset. Lastly, sections \ref{sec:related} and \ref{sec:conclusion} discuss how our method distinguishes itself from related approaches as well as practical limitations of our method.

@@

% (1 page)
\subsection{An idea from Perturbation Theory}
Perturbation Theory comprises methods for finding an approximate solution to a problem by starting from the exact solution of a related, simpler problem and adding successive ``perturbations'' to the system. It is a diverse set of tools used to reason about complex dynamical systems often used in physics and quantum chemistry \cite{hirschfelder-dev-perturb}. We refer the reader to \cite{intro-pertub-theory} for a detailed treatment of the topic.

\begin{definition}[\textbf{Near-idempotent to order $n$}]
    Let the matrix $\vP \in \mathbb{R}^{m \times m}$ satisfy $\vP = \vP^2$. Let ${\vD \in \mathbb{R}^{m \times m}}$ be arbitrary (\textit{e.g.}, noise) where there exists some $n \in \mathbb{N}$ such that $\vD^{n+1}$ has coefficients with absolute value below $\epsilon \ll 1$. We then say that $\vK = \vP + \vD$ is \textbf{near-idempotent to order $n$}.
    \label{def:near-idem}
\end{definition}

Using definition \ref{def:near-idem} we may define the following ansatz in terms of a \textit{near-idempotent} $\vK$:
%
\begin{align}
    \vK' = \alpha_1 \vK + \alpha_2 \vK^2 + \cdots + \alpha_j \vK^j.
    \label{eq:ansatz}
\end{align}
%
This poses $\vK'$ as the linear combination of higher orders of near-idempotent matrices. If we further constrain $(\vK')^2 - \vK' = \bm{0}$, the result is a linear program in variables $\alpha_i$. Importantly, for all equations of the program, any term in which $\vD$ appears at least $n+1$ times can be considered ``negligible'' and ignored. This simplification vastly reduces the problem and allows approximate solutions to the linear program. The coefficients $\alpha_i$ can be thought of as parameterizing a projection $g$ such that $\vK' = g(\vK)$ for an arbitrary near-idempotent $\vK$. The requirement that $\vK'$ be idempotent and that $\vK$ is only near-idempotent implies that a solution $g$ is a projection onto the manifold of idempotent matrices; we call $g$ an \textbf{idempotent corrector} as it must ``make $\vK$ idempotent''.

@@

In the case when $n=1$ we consider $\vD^2 \approx \bm{0}$ and the expression $(\vK')^2 - \vK' = \bm{0}$ can be expanded and reduced by recursively applying the assumptions
%
\begin{align*}
    \vD^2 \approx \bm{0}, \quad \vP^2 = \vP, \quad \vX \vD \vY \vD \vZ \approx \bm{0} \mathrm{~for~all~} \vX,\vY,\vZ
\end{align*}
%
In the case when $j\leq2$, i.e., the ansatz in eq. \ref{eq:ansatz} is $\vK' = \alpha_1 \vK + \alpha_2 \vK^2$, there exists no solutions for $\alpha_i$. When $j=3$, however, there is exactly one solution when $\alpha_1 = 0$, $\alpha_2 = 3$ and $\alpha_3 = -2$, hence $g$ is
%
\begin{align}
    g(\vK) = 3\vK^2 - 2 \vK^3
    \label{eq:}
\end{align}
%
For $j>3$ there exists families of solutions (see Appendix \ref{app:solutions}), but we consider primarily the case when $j=3$ as this requires fewer higher-order terms of $\vK$.

% (1 page)
\subsection{Fixed Points and Stability Analysis}
Undoubtedly, a required property of any idempotent corrector $g$ is that every idempotent matrix is a fixed point, but it may also be desirable to find if any non-idempotent matrices are fixed points. Concretely, we wish to characterize solutions to $\vK = 3 \vK^2 - 2 \vK^3$.

@@

In general, we place no restrictions on the matrix $\vK \in \mathbb{R}^{m \times m}$, hence it might not be diagonalizable directly. It is well known, however, that for every square matrix $\vK$ there exists an invertible matrix $\vP$ such that
\begin{align*}
    \vK = \vP \vJ \vP^{-1}
\end{align*}
where $\vJ \in \mathbb{C}^{m \times m}$ is the Jordan Normal form \cite{jordan-form} of $\vK \in \mathbb{R}^{m \times m}$. From this the dual problem,
\begin{align*}
    \vJ = 3 \vJ^2 - 2 \vJ^3,
\end{align*}
can be constructed. The block-diagonal structure of $\vJ$ imposes up to four equations per eigenvalue of $\vK$ (see Appendix \ref{app:jordan}):
\begin{align}
    \lambda & = 3\lambda^2 - 2\lambda^3 & \label{eq:cond1}                                   \\
    1       & = 6\lambda - 6\lambda^2   & \text{Only when $d_\lambda\geq2$.}\label{eq:cond2} \\
    0       & = 3 - 6\lambda            & \text{Only when $d_\lambda\geq3$.}\label{eq:cond3} \\
    0       & = 0 - 2                   & \text{Only when $d_\lambda\geq4$.}\label{eq:cond4}
\end{align}
where $d_\lambda$ denote the algebraic multiplicity of eigenvalue $\lambda$ of $\vK$. Clearly, this system of equations is inconsistent when $d_{\lambda} \geq 2$, hence geometric multiplicity of every eigenvalue must also be exactly 1. Together this implies that $\vJ$ is diagonalizable for any fixed point $\vK$. Furthermore, the solutions which satisfy only eq. \ref{eq:cond1} are
\begin{align*}
    \lambda \in \{0, 0.5, 1\}
\end{align*}
hence any fixed point of $\vK = 3 \vK^2 - 2 \vK^3$ must have eigenvalues in this set. Consequently, all idempotent matrices are fixed points, but there exists also non-idempotent fixed points.

Although the initial derivation of $g(\vK) = 3 \vK^2 - 2 \vK^3$ relies on $\vK$ being near-idempotent to the first order, we consider more generally the behaviour of $g$ around the fixed points when applied repeatedly as a recurrence relation. Note first that $h(\lambda) = 3\lambda^2 - 2\lambda^3$ has derivative $h'(\lambda) = 6\lambda - 6\lambda^2$, and so for each fixed point of $g$ we have
\begin{align*}
    h'(0) = 0, \quad h'(0.5) = 1.5, \quad h'(1) = 0.
\end{align*}

Since $|h'(\lambda)| < 1$ for $\lambda \in \{0, 1\}$ these points are attracting whilst $|h'(\lambda)| > 1$ for $\lambda = 0.5$, hence this point is repelling. In other words, if the idempotent corrector $g$m applied as a recurrence relation on $\vK$, converges at some point $\vK'$, then $\vK'$ will be approximately idempotent unless $\vK$ has an eigenvalue of exactly $0.5$.

Furthermore, figure \ref{fig:fractal} shows the result of applying the idempotent corrector recursively 10 times for each point on the complex plane. The attracting regions around $0$ and $1$ are large, hence any matrix that is ``reasonably close'' to idempotent will be projected onto a (within machine precision) idempotent matrix.

@@

% (1 page)
\subsection{Deriving a Training Scheme}
\label{sec:method-scheme}
Gradient-based optimization techniques use the gradient of a non-convex loss function as the directional information used to update the hypothesis at each time step. This highlights a core difference between our approach and conventional gradient-based approaches, since the recurrence relation derived above (and shown in figure \ref{fig:plot-g}) exactly describes the ``direction'' to move in to reduce idempotent error. Out method only \textit{evaluates} $g$ -- finding its derivative is unimportant.

Consider a neural network $f_{\vtheta}: \mathbb{R}^m \to \mathbb{R}^m$ together with its application to input $\vx \in \mathbb{R}^m$, denoted $\vy = f_{\vtheta}(\vx)$. We might then consider the recurrence relation above in the following form:
%
\begin{align*}
    \vy' = 3f_{\vtheta}(\vy) - 2f_{\vtheta}(f_{\vtheta}(\vy))
\end{align*}
%
This describes a desired change in the output of the network which we denote ${\Delta f_{\vtheta}(\vx) = \vy' - \vy}$. In other words, $\Delta f_{\vtheta}(\vx)$ describes the desired change in $\vy$ which moves $\vy$ towards an idempotent projection much in the same way that the quantity $\pd{(-\mathcal{L}(\vy))}{\vy}$ describes the direction which reduces the idempotent loss function $\mathcal{L}(\vy)$ in eq. \ref{eq:idem-loss}. In this work, we therefore define
%
\begin{align}
    \pd{(-\mathcal{L}(\vy))}{\vy} \equiv \Delta f_{\vtheta}(\vx)
    \label{eq:substitution}
\end{align}
%
as an alternative quantity to the traditional, analytical solution to $\pd{(-\mathcal{L}(\vy))}{\vy}$.

To complete the scheme, we consider how a change in the output $\vy$ can be propagated to a change in the parameters $\vtheta$ of $f_{\vtheta}$. This, however, is a straightforward application of the chain rule as it is calculated conventionally in backpropagation.

In practice, the definition \ref{eq:substitution} can be implemented in common machine learning frameworks, such as Jax\footnote{https://github.com/jax-ml/jax} and PyTorch\footnote{https://pytorch.org} as a user-defined automatic differentiation rule (see Appendix \ref{app:autodiff-rule}).

% (3 pages)
\section{Experimental Results}
\label{sec:experiment}
%\textit{On experimental networks we give line graphs comparing absolute error at varying learning rates. We also give line graphs for LRs over many epochs. Takeaway message is that for deeper/wider networks we outperform ordinary autodiff by a lot.}
% Outline differences in training scheme.
% Error across LR for B3&B4
% Error across Runs for B3&B4
% (PCA analysis graphs)
To evaluate the training scheme suggested in section \ref{sec:method} we compare relative performance between the two methods: ``Ordinary Backpropagation'' with the quantity $\pd{(-\mathcal{L}(\vy))}{\vy}$ resolved at runtime by automatic differentiation, and ``Modified Backpropagation'' with the modified backpropagation rule for $\pd{(-\mathcal{L}(\vy))}{\vy}$. To demonstrate the flexibility of the approach, we report results for four diverse MLP-style networks, as described in table \ref{tab:networks}.

@@

\subsection{Qualitative differences}
\label{sec:experiment-qual}
%\textit{TODO: Include a section on qualitative differences in behaviour of modified and ordinary backpropagation. Does one method travel a different route in the loss landscape? This small section is to be suggestive of distinguishing features of the new method.}
%
In this section we present suggestive evidence that Modified Backpropagation searches the solution space differently from Ordinary Backpropagation.
%
\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\columnwidth]{./resources/pca_b2.png}
    \caption{Representative projections of the optimizer trajectories over 2500 epochs of either algorithm on the B2 model at optimal learning rates (fig. \ref{fig:avg-abs-err}). Total variance captured is $>97.8\%$ with cosine similarity of PC1 and PC2 less than $1.0\times10^{-6}$. Optimizer trajectory of Modified Backpropagation deviates significantly from Ordinary Backpropagation.}
    \label{fig:pca-b2}
\end{figure}

For purposes of visualization, we employ the method of \cite{li-visualizing} to compare the optimizer trajectories of Modified Backpropagation and Ordinary Backpropagation. Concretely, we train a copy of the same network with either algorithm and record model parameters $\vtheta_t^{\mathrm{MB}}$ and $\vtheta_t^{\mathrm{OB}}$ at epoch $t$. A PCA analysis is then performed over the relative change in parameters from $\vtheta_0^{\mathrm{MB}}$ and $\vtheta_0^{\mathrm{OB}}$ (which are identical), from which we select the two most explanatory directions. An example is shown in figure \ref{fig:pca-b2} (and Appendix \ref{app:pca}).

Qualitative evaluation show that Modified and Ordinary Backpropagation often differ significantly in projected trajectories across the two most explanatory directions, but this is not always the case (e.g., B4 in figure \ref{fig:big-pca}). Additionally, optimization trajectories for Modified Backpropagation can be explained by projection onto two direction with more than $90\%$ variance explained, indicating that it exhibits the same behaviour as Ordinary Backpropagation which has previously been shown to largely operate in low-dimensional subspaces \cite{li-visualizing}. One should note, however, that the loss surface is here represented under a dramatic dimensionality reduction which makes further conclusions unreliable.

>>>>

% Wider context -> Idempotent networks
Using neural networks as data augmentation tools is becoming more widespread in areas such as signal processing and generative artificial intelligence. In particular, networks of the form ${f: X \to X}$, mapping data within the same space $X$, are frequently used in image augmentation \cite{lu-image-aug}, video generation \cite{ma-compression, liu-gans}, sorting algorithms \cite{tambouratzis-sorting}, compression algorithms \cite{namphol-compression, liu-gans}, image denoising \cite{mao-deblurring, ilesanmi-denoising, liu-gans}, and image generation \cite{liu-gans}, among others. While there is a variety of such transformation tasks, some of these can be considered \textit{intrinsically idempotent} in the sense that any function which carries out the transformation is necessarily an idempotent operator. An idempotent operation is one which can be applied iteratively with no effect beyond the first application. For instance, a sorting transformation is intrinsically idempotent since sorting an already sorted data structure is redundant and such a sorting is often deterministic. On the other hand, image transformations might not always be idempotent: rotating an image by 90$^{\circ}$ twice is not the same as rotating it once, for instance. As such, some transformations are intrinsically idempotent and admit \textit{only} idempotent solutions while other tasks are not and admit \textit{no} idempotent solutions. This work, however, is concerned with a class of data transformation tasks which has both idempotent and non-idempotent solutions and where idempotency might be a desirable property. For example, in Section \ref{sec:experiment} we study idempotency in generative networks where it is the formal requirement of one-step inference, but also denoising and image augmentation (\textit{e.g.}, application of effect-filters) are examples of tasks where idempotent solutions may be desirable \cite{mao-deblurring,liu-gans}. Since solutions are not intrinsically idempotent in this class, we explore actively enforcing idempotency as a component of the loss function used in training.

% What is an idempotent network even?
In this paper we are primarily concerned with networks ${f_{\vtheta}: \mathbb{R}^n \to \mathbb{R}^n}$, where $\vtheta$ is a collection of weight parameters. The condition that $f_{\vtheta}$ is idempotent is
%
\begin{equation}
    f_{\vtheta}(\vx) = f_{\vtheta}(f_{\vtheta}(\vx)),
    \label{eq:idem}
\end{equation}
%
for all ${\vx \in \mathbb{R}^n}$. If ${f_{\vtheta}(\vx) = \vW \vx}$ (a single-layer, fully-connected network with no bias and the identity activation function) where ${\vW \in \mathbb{R}^{n \times n}}$ is the weight matrix, then condition (\ref{eq:idem}) reduces to the familiar notion from linear algebra where $\vW = \vW^2$ and eigenvalues of $\vW$ are either 0 or 1. Condition (\ref{eq:idem}) also gives the correct notion for non-linear networks acting as idempotent projections, and can be optimized using a simple mean-squared error loss
%
\begin{align}
    \mathcal{L}_\text{idem}(\vx) = \frac{1}{m} \sum_{i = 1}^m \left(f_{\vtheta}(f_{\vtheta}(\vx)) - f_{\vtheta}\big(\vx \big)\right)^2,
    \label{eq:idem-loss}
\end{align}
%
where ${\vx \in \mathbb{R}^{n}}$. As we show in Section \ref{sec:experiment}, minimizing this loss using canonical gradient descent can yield relatively poor improvement in the idempotent loss. Additionally, due to the higher-order application of $f_{\vtheta}$ the number of terms in the gradient ${\nabla_{\vtheta} \mathcal{L}_{\text{idem}}}$ grows exponentially in the number of layers, making the approach computationally expensive for certain architectures.

% What are we proposing in this paper?
In this work, we propose an alternative method for training neural networks to satisfy condition (\ref{eq:idem}). Using ideas from Perturbation Theory \cite{intro-pertub-theory} we derive a function $g$ which solves ${\vK' = g(\vK)}$ such that if ${\vK \in \mathbb{R}^{n \times n}}$ is an ``almost'' idempotent matrix, then ${\vK' \in \mathbb{R}^{n \times n}}$ is perfectly idempotent (\textit{i.e.}, ${(\vK')^2 = \vK'}$). In this work, we focus on one such function:
%
\begin{equation}
    g(\vK) = 3 \vK^2 - 2 \vK^3.
    \label{eq:g}
\end{equation}

Although we assume $\vK$ is close to idempotent, we show that in practice $g$ can be used to derive matrices which are within machine precision of perfect idempotency even when the input matrix $\vK$ is relatively far from idempotent. At a high level, this process is based on a recurrence relation ${\vK' = \vK + \gamma(g(\vK) - \vK)}$, taking small $\gamma$-sized steps in the direction of $g(\vK)$. While this recurrence relation derives idempotent matrices -- and can therefore be used to train single-layer networks with identity activations to be idempotent -- we also give a more general application of Eq. (\ref{eq:g}) as a modification of the backpropagation algorithm, yielding an architecture agnostic and efficient algorithm for finding idempotent networks.

% Outline of each section of the paper
In Section \ref{sec:method-idea} we give a detailed description of the method used to derive Eq. (\ref{eq:g}) and alternative solutions. We also show that while there exists non-idempotent fixed points to Eq. (\ref{eq:g}), these points are repelling under the recurrence relation ${\vK' = \vK + \gamma(g(\vK) - \vK)}$ for $0 \leq \gamma \leq 1$, giving credence to the use of such a recurrence relation in practice. Finally, in Section \ref{sec:method-scheme} we derive a full training scheme for training arbitrary neural network architectures of the form ${f_{\vtheta}: \mathbb{R}^n \to \mathbb{R}^n}$. In Section \ref{sec:experiment}, we present experimental data for a variety of fully-connected network architectures, showing that our method outperforms ordinary backpropagation under varied conditions. We also replicate the results of \citealt{shocher-ign} by applying our method on a U-net style DCGAN network to successfully create a generative network for the MNIST dataset. Lastly, Sections \ref{sec:related} and \ref{sec:conclusion} discuss how our method distinguishes itself from related approaches as well as future work.

@@

% (1 page)
\subsection{An idea from Perturbation Theory}
\label{sec:method-idea}
Perturbation Theory comprises methods for finding an approximate solution to a problem by starting from the exact solution of a related, simpler problem and adding successive ``perturbations'' to the system. It is a diverse set of tools used to reason about complex dynamical systems often used in physics and quantum chemistry \cite{hirschfelder-dev-perturb}. We refer the reader to \cite{intro-pertub-theory} for a detailed treatment of the topic.

We first define the central term \textit{near-idempotent}, used throughout:

\begin{definition}[\textbf{Near-idempotent to order $n$}]
    Let the matrix ${\vP \in \mathbb{R}^{m \times m}}$ satisfy ${\vP = \vP^2}$. Let ${\vD \in \mathbb{R}^{m \times m}}$ be arbitrary (\textit{e.g.}, noise) where there exists some ${n \in \mathbb{N}}$ such that $\vD^{n+1}$ has coefficients with absolute value below $\epsilon \ll 1$. We say that ${\vK = \vP + \vD}$ is \textbf{near-idempotent to order $n$}.
    \label{def:near-idem}
\end{definition}

Using definition \ref{def:near-idem} we may define the following ansatz in terms of a near-idempotent $\vK$:
%
\begin{align}
    \vK' = \alpha_1 \vK + \alpha_2 \vK^2 + \cdots + \alpha_j \vK^j.
    \label{eq:ansatz}
\end{align}
%
This poses $\vK'$ as the linear combination of higher orders of near-idempotent matrices. If we further constrain ${(\vK')^2 - \vK' = \bm{0}}$, the result is a linear program in variables $\alpha_i$. Importantly, for all equations of the program, any term in which $\vD$ appears at least $n+1$ times can be considered ``negligible'' and ignored. This simplification vastly reduces the problem and allows approximate solutions to the linear program. The coefficients $\alpha_i$ can be thought of as parameterizing a projection $g$ such that ${\vK' = g(\vK)}$ for an arbitrary near-idempotent $\vK$. The requirement that $\vK'$ be idempotent and that $\vK$ is only near-idempotent implies that a solution $g$ is a projection onto the manifold of idempotent matrices; we call $g$ an \textbf{idempotent corrector} as it must ``make $\vK$ idempotent''.

@@

In the case when $n=1$ we consider ${\vD^2 \approx \bm{0}}$ and the expression ${(\vK')^2 - \vK' = \bm{0}}$ can be expanded and reduced by recursively applying the assumptions
%
\begin{align}
    \vD^2 \approx \bm{0}, \quad \vP^2 = \vP, \quad \vX \vD \vY \vD \vZ \approx \bm{0},
\end{align}
%
for all $\vX,\vY,\vZ$ matrices. When $j\leq2$, \textit{i.e.}, the ansatz in Eq. (\ref{eq:ansatz}) is ${\vK' = \alpha_1 \vK + \alpha_2 \vK^2}$, there exists no solutions for $\alpha_i$. When $j=3$, however, there is exactly one solution when ${\alpha_1 = 0}$, ${\alpha_2 = 3}$ and ${\alpha_3 = -2}$, thus $g$ is as defined in Eq. (\ref{eq:g}). For $j>3$ there exists families of solutions (see Appendix \ref{app:solutions}), but we consider primarily the case when $j=3$ as this requires fewer higher-order terms of $\vK$ and is therefore generally less costly to compute.

% (1 page)
\subsection{Fixed Points and Stability Analysis}
\label{sec:method-fixed-stab}
Undoubtedly, a required property of any idempotent corrector $g$ is that every idempotent matrix is a fixed point, but it may also be desirable to find if any non-idempotent matrices are fixed points. Concretely, we wish to characterize solutions to ${\vK = 3 \vK^2 - 2 \vK^3}$.

@@

In general, we place no restrictions on the matrix ${\vK \in \mathbb{R}^{m \times m}}$. In particular, it might not be directly diagonalizable. It is well known, however, that for every square matrix $\vK$ there exists an invertible matrix $\vP$ such that
\begin{align}
    \vK = \vP \vJ \vP^{-1}
\end{align}
where $\vJ \in \mathbb{C}^{m \times m}$ is the Jordan normal form \cite{jordan-form} of ${\vK \in \mathbb{R}^{m \times m}}$. From this the dual problem,
\begin{align}
    \vJ = 3 \vJ^2 - 2 \vJ^3,
\end{align}
can be constructed. The block-diagonal structure of $\vJ$ imposes up to four equations per eigenvalue of $\vK$ (see Appendix \ref{app:jordan}):
\begin{align}
    \lambda & = 3\lambda^2 - 2\lambda^3 & \label{eq:cond1}                                   \\
    1       & = 6\lambda - 6\lambda^2   & \text{Only when $d_\lambda\geq2$.}\label{eq:cond2} \\
    0       & = 3 - 6\lambda            & \text{Only when $d_\lambda\geq3$.}\label{eq:cond3} \\
    0       & = 0 - 2                   & \text{Only when $d_\lambda\geq4$.}\label{eq:cond4}
\end{align}
where $d_\lambda$ denotes the algebraic multiplicity of eigenvalue $\lambda$ of $\vK$. Clearly, this system of equations is inconsistent when ${d_{\lambda} \geq 2}$, hence geometric multiplicity of every eigenvalue must also be exactly 1. Together this implies that $\vJ$ is diagonalizable for any fixed point $\vK$. Furthermore, the solutions which satisfy only Eq. (\ref{eq:cond1}) are:
%
\begin{align}
    \lambda \in \{0, 0.5, 1\}.
\end{align}

Therefore, any fixed point of ${\vK = 3 \vK^2 - 2 \vK^3}$ must have eigenvalues in this set. Consequently, all idempotent matrices are fixed points, but there exists also non-idempotent fixed points.

Although the initial derivation of ${g(\vK) = 3 \vK^2 - 2 \vK^3}$ relies on $\vK$ being near-idempotent to the first order, we consider more generally the behaviour of $g$ around the fixed points when applied repeatedly as a recurrence relation. Let ${h(\lambda) = 3\lambda^2 - 2\lambda^3}$ and observe its derivative ${h'(\lambda) = 6\lambda - 6\lambda^2}$. Then, for each fixed point of $g$ we have
\begin{align}
    h'(0) = 0, \quad h'(0.5) = 1.5, \quad h'(1) = 0.
\end{align}

Since ${|h'(\lambda)| < 1}$ for ${\lambda \in \{0, 1\}}$ these points are attracting whilst ${|h'(\lambda)| > 1}$ for ${\lambda = 0.5}$, thus this point is repelling. In other words, if the idempotent corrector $g$, applied as a recurrence relation on $\vK$, converges at some point $\vK'$, then $\vK'$ will be approximately idempotent unless $\vK$ has an eigenvalue of exactly $0.5$.

Furthermore, Figure \ref{fig:fractal} shows the result of applying the idempotent corrector recursively 10 times for each point on the complex plane. The attracting regions around $0$ and $1$ are large, hence any matrix that is ``reasonably close'' to idempotent will be projected onto a (within machine precision) idempotent matrix.

@@

Whilst this only technically applies in the linearized setting, we propose to also apply the method in non-linear settings using the recurrence relation
%
\begin{align}
    \vK' = \vK + \gamma(g(\vK) - \vK)
\end{align}
%
for $0 \leq \gamma \leq 1$. This has the effect of taking small $\gamma$-sized steps in the direction of $g$ at every time point.

% (1 page)
\subsection{Deriving a Training Scheme}
\label{sec:method-scheme}
Gradient-based optimization techniques use the gradient of an often non-convex loss function as the directional information used to update the hypothesis at each time step. This highlights a core difference between our approach and conventional gradient-based approaches, since the recurrence relation derived above (and shown in Figure \ref{fig:plot-g}) exactly describes the ``direction'' to move in to reduce idempotent error. Our method need only \textit{evaluate} $g$ -- finding its derivative is unimportant.

Consider a neural network ${f_{\vtheta}: \mathbb{R}^m \to \mathbb{R}^m}$ together with its application to input ${\vx \in \mathbb{R}^m}$, denoted ${\vy = f_{\vtheta}(\vx)}$. We might then consider the recurrence relation in Eq. (\ref{eq:g}) in the following form:
%
\begin{align}
    \vy' = 3f_{\vtheta}(\vy) - 2f_{\vtheta}(f_{\vtheta}(\vy))
\end{align}
%
This describes a desired change in the output of the network which we denote ${\Delta f_{\vtheta}(\vx) = \vy' - \vy}$. In other words, ${\Delta f_{\vtheta}(\vx)}$ describes the change in $\vy$ which moves $\vy$ towards an idempotent projection much in the same way that the quantity $\pd{(-\mathcal{L}_{\mathrm{idem}}(\vy))}{\vy}$ describes the direction which reduces the idempotent loss function $\mathcal{L}_{\mathrm{idem}}(\vy)$ in Eq. (\ref{eq:idem-loss}). A central idea presented in this work is therefore the definition
%
\begin{align}
    \pd{(-\mathcal{L}(\vy))}{\vy} \equiv \Delta f_{\vtheta}(\vx)
    \label{eq:substitution}
\end{align}
%
as an alternative quantity to the traditional, analytical solution to $\pd{(-\mathcal{L}(\vy))}{\vy}$. To complete the scheme, we consider how a change in the output $\vy$ can be propagated to a change in the parameters $\vtheta$ of $f_{\vtheta}$. This, however, is a straightforward application of the chain rule as it is calculated conventionally in backpropagation. We use the term ``\textbf{Modified Backpropagation}'' to refer to the canonical backpropagation algorithm with the rule (\ref{eq:substitution}) applied appropriately when computing gradients.

In practice, the definition (\ref{eq:substitution}) can be implemented in common machine learning frameworks, such as Jax\footnote{https://github.com/jax-ml/jax} and PyTorch\footnote{https://pytorch.org} as a user-defined automatic differentiation rule (see Appendix \ref{app:autodiff-rule}).

% (3 pages)
\section{Experimental Results}
\label{sec:experiment}
%\textit{On experimental networks we give line graphs comparing absolute error at varying learning rates. We also give line graphs for LRs over many epochs. Takeaway message is that for deeper/wider networks we outperform ordinary autodiff by a lot.}
% Outline differences in training scheme.
% Error across LR for B3&B4
% Error across Runs for B3&B4
% (PCA analysis graphs)
To evaluate the training scheme suggested in Section \ref{sec:method-scheme} we compare relative performance between the two methods: ``Ordinary Backpropagation'' with the quantity $\pd{(-\mathcal{L}(\vy))}{\vy}$ resolved at runtime by automatic differentiation, and ``Modified Backpropagation'' with the modified backpropagation rule for $\pd{(-\mathcal{L}(\vy))}{\vy}$. To demonstrate the flexibility of the approach, we report results for four diverse MLP-style networks, as described in table \ref{tab:networks}.

@@

\subsection{Qualitative differences}
\label{sec:experiment-qual}
%\textit{TODO: Include a section on qualitative differences in behaviour of modified and ordinary backpropagation. Does one method travel a different route in the loss landscape? This small section is to be suggestive of distinguishing features of the new method.}
%
In this section we present suggestive evidence that Modified Backpropagation searches the solution space differently from Ordinary Backpropagation.
%
\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\columnwidth]{./resources/pca_b2.png}
    \caption{Representative projections of the optimizer trajectories over 2500 epochs of either algorithm on the B2 model at optimal learning rates (Figure \ref{fig:avg-abs-err}). Total variance captured is $>97.8\%$ with cosine similarity of PC1 and PC2 less than $1.0\times10^{-6}$. Optimizer trajectory of Modified Backpropagation deviates significantly from Ordinary Backpropagation.}
    \label{fig:pca-b2}
\end{figure}

For purposes of visualization, we employ the methods of \citealt{li-visualizing} to compare the optimizer trajectories of Modified Backpropagation and Ordinary Backpropagation. Concretely, we train a copy of the same network with either algorithm and record model parameters $\vtheta_t^{\mathrm{MB}}$ and $\vtheta_t^{\mathrm{OB}}$ at epoch $t$. A PCA analysis is then performed over the relative change in parameters from $\vtheta_0^{\mathrm{MB}}$ and $\vtheta_0^{\mathrm{OB}}$ (which are identical), from which we select the two most explanatory directions. Lastly, the loss landscape and trajectory paths $\vtheta_t^{\mathrm{MB}}$ and $\vtheta_t^{\mathrm{OB}}$ are projected onto the selected dimensions. An example is shown in Figure \ref{fig:pca-b2} (and Appendix \ref{app:pca}).
